Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-04-30 manu-draft.qmd
more revisions

Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.04 $^\circ$C per year and salinity decreasing by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the last fifty years. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay. Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model.

@@

Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`). Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`). Using this model, mean precipitation in 1975 was `r prpstr2` mm, whereas current mean precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm. Notably, trends for the dry season (October through May) were not significant (`r drypvl`), except April which had a significantly increasing trend (`r aprpvl`, $R^2$ = `r aprrsq`). The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.5 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had an approximate increase of 50 days each year for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022). The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significantly increasing model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.5 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were more than twice those in HB. Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5). Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors. Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.

@@

This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design. An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future. Simple regression models provided weak, but partially-supporting evidence, that these changes can be linked to recent seagrass losses since 2016. Our models suggested that rising temperatures and decreasing salinities have had additive rather than multiplicative effects on seagrass, as evidenced by lack of significant interactions in models involving both stressors. Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges. These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales. Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress.

>>>>

Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.04 $^\circ$C per year and salinity decreasing by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the last fifty years. Additionally, the number of days when temperature was above 30 $^\circ$C or salinity was below 25 ppt has increased on average across all bay segments by 48 and 37 days, respectively, since 1975. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay. Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions and continuous monitoring data to support ecosystem health into the future.

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries. As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).

@@

Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`). Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`), suggesting precipitation increases were driven by the earlier months (June - August). Using this model, mean precipitation in 1975 was `r prpstr2` mm, whereas current mean precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm. Notably, trends for the dry season (October through May) were not significant (`r drypvl`), except April which had a significantly increasing trend (`r aprpvl`, $R^2$ = `r aprrsq`). The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.5 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record were similar between bay segments, with an average increase of 48 days across the bay segments for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022). The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significantly increasing model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.5 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were more than twice those in HB. Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5). Overall, OTB showed the largest increase in the number of days each year when salinity was below a threshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022). Across all bay segments, the average increase in the number of days salinity was below 25 ppt from 1975 to 2022 was 36 days, although a distinct gradient towards the mouth of the bay was observed.

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This is especially true for *H. wrightii* that had the greatest changes over the period of record and is tolerant of a wide range of salinity. However, this does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors. Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.

@@

This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design. An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future. Simple regression models provided weak, but partially-supporting evidence, that these changes can be linked to recent seagrass losses since 2016. Our models suggested that rising temperatures and decreasing salinities have had additive rather than multiplicative effects on seagrass, as evidenced by lack of significant interactions in models involving both stressors. Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges. These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales. Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see <https://tbep-tech.github.io/otb-temp/tempeval>). Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress.
